---
title: "SwathFinding"
author: "wkumler"
date: "August 6, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MSnbase)
library(dplyr)
```

Mass values in a mass spectrometer are collected in "swaths". Finding these
swaths of data collection is important when squaring MS data because then
each data point can be assigned a single mz/rt value across different files.

However, these swaths are not equally spaced, but instead separated by a
distance proportional to the mass of the molecule. Further, this separation
is not linear with mass but instead has a gentle curve that appears to be
best matched by second-order polynomials.

Swaths can be collapsed into peaks by histogramming the number of values found
at a given m/z value. Histogram bins with a high count will be closer to the
center of the swath, while histogram bins with low or zero count are found
between swaths, as seen in the figure below.

```{r histogramming, warning=F, message=F}
#SwathFindingMiniData was created by filtering
# Z:/1_QEdata/Will/RectangulaRdata/positive/190715_Blk_KM1906U14-Blk_C.mzML 
# to masses less than 120 m/z and retention times less than 300 s

simple_data <- readMSData("SwathFindingMiniData.mzML", centroided. = F, msLevel. = 1)

layout(matrix(c(1,1,2,1,1,2), byrow = T, nrow = 2))
par(mar=c(4.1, 4.1, 0, 0))
x <- simple_data %>%
  filterMz(c(90.054, 90.058)) %>%
  as("data.frame")
plot(x$rt, x$mz, main = "", pch = 19, xlab = "Retention time", ylab = "m/z",
     col = lattice:::level.colors(x$i, col.regions = topo.colors, 
                                  at = lattice::do.breaks(range(x$i), nint = 256)))
grid()
par(mar=c(4.1, 0, 0, 0))
barplot(hist(x$mz, breaks = 1000, plot=F)$counts, 
        axes = T, space = 0, horiz=TRUE, xlab="Counts", ylab="")
grid()
```

The swaths are fairly regular, so we'd like to find the "typical" distance
between swaths and use that to slice our data into nice bins, each bin
including one m/z swath. To do that, we apply some peakfinding software from
MALDIquant to find the local maximum of each peak in the histogram and report
the distance between the local maxima.

```{r peakpicking}
mz_raw <- unlist(mz(simple_data))

segment_size <- 7
lowest_mz <- floor(min(mz_raw))-floor(min(mz_raw))%%segment_size
highest_mz <- floor(max(mz_raw))+floor(max(mz_raw))%%segment_size
bin_minima <- seq(lowest_mz, highest_mz, segment_size)

getMeds <- function(bin_min, mz_raw){
  raw_i <- mz_raw[mz_raw>(bin_min)&mz_raw<(bin_min+segment_size)]
  span <- round(bin_min/segment_size)
  peak_mids <- list()
  for(j in seq(0, segment_size-0.1, 0.1)){
    raw_bin_j <- raw_i[raw_i>(bin_min+j)&raw_i<(bin_min+j+0.1)]
    if(length(raw_bin_j)>1){
      hist_bin_j <- hist(raw_bin_j, breaks = 10000, plot = F)
      peak_idxs <- MALDIquant:::.localMaxima(hist_bin_j$counts, span)
      peak_mids[[(j+0.1)*10]] <- hist_bin_j$mids[peak_idxs]
    }
  }
  return(peak_mids)
}

mids <- unlist(lapply(bin_minima, getMeds, mz_raw))

fit_df <- data.frame(mass=mids[-1], diff_y=diff(mids))
fit_df_smallnums <- filter(fit_df, diff_y<0.001)
layout(matrix(c(1,1,2,1,1,2), byrow = T, nrow = 2))
plot(fit_df$mass, fit_df$diff_y, las=1,
     ylab="Distance between swaths (Da)", xlab="Molecule mass")
plot(fit_df_smallnums$mass, fit_df_smallnums$diff_y, las=1,
     ylab="Distance between swaths (Da)", xlab="Molecule mass")
```

This plot is more or less exactly what we expected. Remember, we're looking for
a positive correlation between molecular mass and distance between swaths.

The left plot shows the raw
differences between swaths, but this is hard to see patterns in because a few
outliers throw the y-axis out of whack. On the right, I've zoomed in to just
swath differences less than 0.001 Da - still showing ~95% of the data. Here,
we can see the trends that we were looking for. The thick line along the bottom
is the "true" signal, the typical distance between swaths. The fainter upper
trends occur when a peak is missing in the data - for example, in regions where
molecules themselves are few and far between. These show the same trend as the
actual data, but have a slope double, triple, etc. the base signal as one, two, 
or more peaks are absent.

What we'd like to do is eliminate the false signal here and run a linear
regression on the true data points. I'm going to do this semi-manually by
fitting quadratic equations to three points along the curve, then calculating
the coefficients via some linear algebra magic.

```{r quadratics}
curveMaker <- function(x, formula){formula[1]*x**2+formula[2]*x+formula[3]}

test_points <- data.frame(x=c(60, 90, 120), 
                           y_upper=c(0.00020, 0.00035, 0.00053),
                           y_lower=c(0.00007, 0.00011, 0.00021))

form_mat <- rbind(c(test_points$x[1]**2, test_points$x[1], 1), 
                 c(test_points$x[2]**2, test_points$x[2], 1),
                 c(test_points$x[3]**2, test_points$x[3], 1))
upper_formula <- solve(form_mat, test_points$y_upper)
upper_bound <- curveMaker(fit_df_smallnums$mass, upper_formula)
lower_formula <- solve(form_mat, test_points$y_lower)
lower_bound <- curveMaker(fit_df_smallnums$mass, lower_formula)

par(mfrow=c(1,2))

plot(fit_df_smallnums$mass, fit_df_smallnums$diff_y,
     ylab="Distance between swaths (Da)", xlab="Molecule mass")
points(test_points$x, test_points$y_upper, col="red", lwd=2, cex=2)
points(test_points$x, test_points$y_lower, col="red", lwd=2, cex=2)
points(fit_df_smallnums$mass, upper_bound, col="red", cex=0.2)
points(fit_df_smallnums$mass, lower_bound, col="red", cex=0.2)

good_vals <- filter(fit_df_smallnums, diff_y<upper_bound&diff_y>lower_bound)
plot(good_vals, ylim=c(0, 0.001),
     ylab="Distance between swaths (Da)", xlab="Molecule mass")
```

We probably could fit the model a lot tighter and remove more noise, but that's 
what the linear model is designed to do - all that we were hoping to accomplish
with this step was removing the signal from the missing peak differences.

Now let's fit the linear model to our pretty data:

```{r linearize}



```